Extended Data

### 2022-03-26 doc/manuscript/Dow_manuscript_Nature.Rmd
revise intro paragraph on modeling
#130

In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;<!--@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;-->@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
Based on these observations, current terrestrial ecosystem models effectively assume that longer growing seasons contribute positively to annual net CO~2~ uptake (i.e., net ecosystem exchange, NEE)[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012], as occurs when additional C is allocated to woody growth and therefore resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014].
However, model representation of C allocation to stem growth does not capture known decoupling of stem growth from photosynthate production[@delpierre_temperate_2016;@delpierre_wood_2016;@fatichi_moving_2014].
As a result, the consequences of rising spring temperatures on stem growth -- or woody aboveground net primary productivity (ANPP~woody~) on the ecosystem level -- may not be accurately represented in models[@delpierre_temperate_2016;@fatichi_moving_2014]. 
Understanding the sensitivity of ANPP~woody~ to spring temperatures is central to predicting the future of the temperate deciduous forest C sink.

Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;<!--@delpierre_chilling_2019;-->@rossi_critical_2008] and increased annual growth in mesic climates[<!--@cook_predicting_1991;-->@babst_twentieth_2019;<!--@anderson-teixeira_joint_2022;--> @gao_earlier_2022]. 
In contrast, we have little evidence as to how stem growth and ANPP~woody~ respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016].
Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020<!--;@stuble_plant_2021;@delgado_differences_2020-->].<!--Most of htese citations are important in other spots, but do we need all of them here?-->
However, earlier initiation of growth would not necessarily translate to earlier, faster, or greater stem growth; rather, stem growth is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021;@tumajer_limitation_2022], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], gross primary productivity (GPP)[@delpierre_wood_2016], or NEE[@delpierre_wood_2016]. 
Furthermore, growth of broadleaf deciduous trees may be sink-saturated[@fatichi_moving_2014], such that longer growing seasons with more C fixation do not necessarily augment growth[@delpierre_wood_2016;@zani_increased_2020;@zohner_how_2021].
Tree-ring records, which can be used to examine annual growth but not growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019], with some evidence that warmer springs may have a modest positive effect on growth[@helcoski_growing_2019;@gao_earlier_2022].
These observations do not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake in temperate deciduous forests[@keenan_net_2014;@lu_no_2022].
Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.

@@

<!--conclusions-->
As spring temperatures become increasingly warmer, growing seasons will start earlier. 
However, barring rapid acclimation of temperate deciduous forests to the warming conditions, advancement in the timing of stem growth (Fig. `r schematic_summary`) is unlikely to provide the sustained increase in CO~2~ sequestration in woody biomass and ensuant negative climate change feedback *that is anticipated in most climate forecasting models*[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
Rather, the dominant effect of rising temperatures on forest woody productivity will be a negative effect of high summer temperatures (Fig. `r tree_rings`), which constitutes a positive feedback to climate change.

>>>>

In temperate deciduous forests, spring warming generally lengthens the period over which trees have photosynthetically active leaves[@jeong_phenology_2011;<!--@ibanez_forecasting_2010;@menzel_european_2006;@menzel_growing_1999;-->@lu_no_2022] and that over which the ecosystem is a net CO~2~ sink[@keenan_net_2014;@lu_no_2022].
Based on these observations, current terrestrial ecosystem models represent warm springs and longer growing seasons as contributing positively to annual gross primary productivity (GPP) and net CO~2~ uptake (i.e., net ecosystem exchange, NEE)[@richardson_terrestrial_2012;@ahlstrom_robustness_2012].
However, the long-term persistence of this extra C in the ecosystem, and the associated negative feedback to climate change[@keenan_net_2014], will depend on the extent to which it is allocated woody growth and therefore resides in the ecosystem for decades to centuries[@xue_global_2017;@russell_residence_2014], as opposed to being rapidly released back to the atmosphere through respiration[@jiang_fate_2020;@oishi_warmer_2018;@buermann_widespread_2018].
Model representation of C allocation to stem growth -- or woody aboveground net primary productivity (ANPP~woody~) on the ecosystem level -- does not capture known decoupling of stem growth from photosynthate production[@delpierre_temperate_2016;@delpierre_wood_2016;@fatichi_moving_2014].
As a result, the consequences of rising spring temperatures on stem growth may not be accurately represented in models[@delpierre_temperate_2016;@fatichi_moving_2014]. 
Understanding the sensitivity of ANPP~woody~ to spring temperatures is central to predicting the future of the temperate deciduous forest C sink.

Most studies on tree growth responses to warmer spring temperatures have focused on boreal or temperate conifers, which tend to respond to warmer spring temperatures with an earlier start to growth[@huang_photoperiod_2020;<!--@delpierre_chilling_2019;-->@rossi_critical_2008] and increased annual growth in mesic climates[<!--@cook_predicting_1991;-->@babst_twentieth_2019;<!--@anderson-teixeira_joint_2022;--> @gao_earlier_2022]. 
In contrast, we have little evidence as to how stem growth and ANPP~woody~ respond to warmer spring temperatures in deciduous forests[@savage_coordination_2021;@delpierre_temperate_2016].
Close coordination of budburst and initiation of xylogenesis[@delpierre_temperate_2016] suggests that warm springs should shift the onset of growth earlier alongside observed advances in leaf phenology[@jeong_phenology_2011;@keenan_net_2014;@buermann_widespread_2018;@finzi_carbon_2020<!--;@stuble_plant_2021;@delgado_differences_2020-->].<!--Most of htese citations are important in other spots, but do we need all of them here?-->
However, earlier initiation of growth would not necessarily translate to earlier, faster, or greater stem growth; rather, stem growth is dependent upon environmental conditions on hourly to daily time scales[@zweifel_why_2021;@tumajer_limitation_2022], and annual growth is more closely linked to conditions within the growing season than to growing season length[@etzold_number_2022], GPP[@delpierre_wood_2016], or NEE[@delpierre_wood_2016]. 
Furthermore, growth of broadleaf deciduous trees may be sink-saturated[@fatichi_moving_2014], such that longer growing seasons with more C fixation do not necessarily augment growth[@delpierre_wood_2016;@zani_increased_2020;@zohner_how_2021].
Tree-ring records, which can be used to examine annual growth but not growth seasonality, reveal that growth of temperate deciduous trees tends to be most sensitive to temperature or potential evapotranspiration between late spring and early summer[@dorangeville_drought_2018;@helcoski_growing_2019], with some evidence that warmer springs may have a modest positive effect on growth[@helcoski_growing_2019;@gao_earlier_2022].
These observations do not necessarily align with the finding that warming spring temperatures increase annual forest CO~2~ uptake in temperate deciduous forests[@keenan_net_2014;@lu_no_2022].
Characterizing responses of stem growth to warming spring temperatures is critical to bridging this conceptual disconnect and understanding how forest biomass growth is likely to change as the climate warms.

@@

<!--conclusions-->
As spring temperatures become increasingly warmer, growing seasons will start earlier. 
However, barring rapid acclimation of temperate deciduous forests to the warming conditions, advancement in the timing of stem growth (Fig. `r schematic_summary`) is unlikely to provide a sustained augmentation of CO~2~ sequestration in woody biomass and ensuant negative climate change feedback *that is anticipated in most climate forecasting models*[@buermann_widespread_2018;@zani_increased_2020;@keenan_net_2014;@ahlstrom_robustness_2012].
Rather, the dominant effect of rising temperatures on forest woody productivity will be a negative effect of high summer temperatures (Fig. `r tree_rings`)[@ahlstrom_robustness_2012], which constitutes a positive feedback to climate change.
